---
title: "DSCI 300 Mini  Project 4 - Cluster Analysis"
author: "Nisi Mohan Kuniyil 300321388"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(cache = TRUE)
```



```{r, echo=FALSE}

library(ggpubr)
library(pixmap)
library(ggplot2)
library(dplyr)
library(gapminder)
library(graphics)
library(ggcorrplot)
library(magrittr)
library(scales)
library(tidyverse)
library(hrbrthemes)
library(viridis)
library(dataMaid)
library(countrycode)
library(reshape)
library(reshape2)
library(janitor)
```





# Tidying Data


**Read only first 5 rows from BigTenexpand.csv, since there are two sets of**
**data in single csv.**

```{r}

set.seed(42)
University_49 <- read.csv("BigTenexpand.csv", nrows = 5 )

head(University_49)

```


**Remove the extra column "i..candidates" from the imported dataset University_49** 


```{r}
University_49 <- University_49[, -1]
head(University_49)

```


**Get the transpose of the dataframe "University_49"**



```{r}

University_49 <- data.frame(t(University_49))
head(University_49)

```
**Create a new column name and assign the first column values to this "newcol" **


```{r}


University_49 <- cbind(newcol = rownames(University_49), University_49)
rownames(University_49) <- 1:nrow(University_49)

head(University_49)

```



**Remove the first row from the new dataset "Univeristy_49" **


```{r}

University_49 <- University_49%>%
  row_to_names(row_number = 1)
head(University_49)

```

**There are lot of extra rows got added after clean up, remove those extra rows.**

```{r}
# removing rextra rows

University_49 <- head(University_49, 49)
head(University_49)
```


**Big Ten Schools importing.**

```{r}

BigTenUniversity <- read.csv("BigTenExpand.csv", skip = 7)[,1:2]
head(BigTenUniversity)

```

**The dataset needs to be changed to wide format, in order to do this we need a** 
**unique id for each group. For this creating a new column to use in reshaping.**


```{r}
#adding a dummy vector to get a unique column for long to wide
vector_col <-  vector(mode = "integer", length = 60)
for(i in seq_len(length(vector_col)-1)){
  
  vector_col[i+1] <- floor(i/6)
  
}

vector_col

```



**combining this vector with our datset**

```{r}

#cbind 

BigTenUniversity <- cbind(BigTenUniversity, vector_col)

head(BigTenUniversity)
```
**Using reshape, changing to wide format from long.**


```{r}
#changing to wide format from long
BigTenUniversity <- reshape(BigTenUniversity, timevar = "Big.Ten", 
                            idvar = "vector_col", direction = "wide")
head(BigTenUniversity)
```


**Rename the column names same as the dataset "University_49".**


```{r}
ColumnNames <- c("vector_col","School", "StadiumCapacity","Latitude",
                 "Longitude","Endowment ($000)","Enrollment")

colnames(BigTenUniversity) <- ColumnNames 

```

```{r}

BigTenUniversity
```




**Creating index as school names**


```{r}

BigTenUniversity <- BigTenUniversity[, -1]
rownames(BigTenUniversity) <- NULL
BigTenUniversity <- column_to_rownames(BigTenUniversity, "School")
BigTenUniversity
```
\newpage

1.Using Euclidean distance to measure dissimilarity between observations,
determine which school (in its own cluster of one) that hierarchical clustering 
with complete linkage would recommend integrating into the Big Ten Conference.
That is, which school is the most similar with respect to complete linkage
to the cluster of ten schools that were members of the Big Ten from 1946 
to 1990?




**Take average of BigTen universities and add this as a fake college in**
**the dataset "University_49"**



```{r}
ColumnNames <- c("StadiumCapacity","Latitude","Longitude",
                 "Endowment ($000)","Enrollment")

BigTenUniversity[ColumnNames] <- sapply(BigTenUniversity[ColumnNames],
                                        as.numeric)
sapply(BigTenUniversity, class)

```

**Taking the mean of BigTenUniversity**

```{r}
BigTen_mean <- colMeans(BigTenUniversity)
BigTen_mean
```
**Appending name to it inorder to add that in University_49**

```{r}
BigTen_mean <- append(BigTen_mean, "BigTen", 0)
BigTen_mean
```

**adding a new row BigTen**


```{r}
University_50 <- rbind(University_49, BigTen_mean)
tail(University_50)
```

**converting columns from char to double**

```{r,echo= FALSE}
#converting to numeric, univerisity with 50 colleges
University_50[ColumnNames] <- sapply(University_50[ColumnNames],as.numeric)
tail(University_50)
```

**Converting column "School" to rownames**

```{r}
rownames(University_50) <- NULL
University_50 <- column_to_rownames(University_50, "School")
tail(University_50)
```

**clustering methord to figure out the distance b/w each college and cluster**
**based on the farthest distance.**

```{r}

maximum_dist <- dist(University_50, method = "maximum")
maximum_dist
```

**School closer to the BigTen**

```{r}

maximum_dist <- as.matrix(maximum_dist)
which.min(maximum_dist["BigTen", -50])
min(maximum_dist["BigTen",-50])


```

**We got Missouri as our first school which is similar to Big Ten**



\newpage


2.Add the single school identified in (1) to create a cluster of 11 schools 
representing a hypothetical BigTen Conference. Repeat the calculations to 
identify the school most similar with respect to complete linkage to this
new cluster of 11 schools.



**Adding Missouri to BigTenUniversity dataframe**
```{r}


BigElevenUniversity <- rbind(BigTenUniversity, University_50["Missouri",])
BigElevenUniversity

```
**Take average of BigElevenUniveristy**
```{r}

BigEleven_mean <- colMeans(BigElevenUniversity)
BigEleven_mean

```
```{r}
BigEleven_mean <- append(BigEleven_mean, "BigEleven", 0)
BigEleven_mean
```

**Removing Missouri from University_49 and adding BigEleven**
```{r}
University_48 <- University_49 %>% filter(School != "Missouri")
head(University_48)
```
**Adding BigElevenMean to University_48**

```{r}
University_49 <- rbind(University_48, BigEleven_mean)
rownames(University_49) <- NULL
University_49 <- column_to_rownames(University_49, "School")
tail(University_49)
```
```{r}
University_49[ColumnNames] <- sapply(University_49[ColumnNames], as.numeric)
tail(University_49)
```


```{r}

maximum_dist <- dist(University_49, method = "maximum")
#maximum_dist
```

School closer to the BigTen

```{r}

maximum_dist <- as.matrix(maximum_dist)

#colnames(maximum_dist)

which.min(maximum_dist["BigEleven", -49])
min(maximum_dist["BigEleven",-49])

```

**We got Iowa.State as our second school which is similar to Big Ten**


\newpage

3.Add the school identified in (2) to create a cluster of 12 schools 
representing a hypothetical Big Ten Conference. Repeat the calculations to 
identify the school most similar with respect to complete linkage
to this new cluster of 12 schools.


**Adding Iowa.State to BigElevenUniversity dataframe**
```{r}


BigTwelveUniversity <- rbind(BigElevenUniversity, University_49["Iowa.State",])
BigTwelveUniversity

```

**Finding mean of BigTwelveUniverisity**


```{r}

BigTwelve_mean <- colMeans (BigTwelveUniversity)
BigTwelve_mean

```



```{r}
BigTwelve_mean <- append(BigTwelve_mean, "BigTwelve", 0)
BigTwelve_mean
```


**Removing Iowa.State from University_48**

```{r}
University_47 <- University_48 %>% filter(School != "Iowa.State")
head(University_47)
```
**Adding BigTwelveMean to University_47**

```{r}
University_48 <- rbind(University_47, BigTwelve_mean)
rownames(University_48) <- NULL
University_48 <- column_to_rownames(University_48, "School")
tail(University_48)
```

**Again typecasting to double**

```{r}
University_48[ColumnNames] <- sapply(University_48[ColumnNames], as.numeric)
head(University_48)
```


```{r}

maximum_dist <- dist(University_48, method = "maximum")
#maximum_dist
```

School closer to the BigTen

```{r}

maximum_dist <- as.matrix(maximum_dist)

#colnames(maximum_dist)

which.min(maximum_dist["BigTwelve", -48])
min(maximum_dist["BigTwelve",-48])

```

**We got Kansas as our third school.**


\newpage


4.Add the school identified in (3) to create a cluster of 13 schools 
representing a hypothetical Big Ten Conference. Repeat the calculations 
to identify the school most similar with respect to complete linkage
to this new cluster of 13 schools. Add this school to create a 14-school cluster.

**Adding Kansas to BigTwelveUniversity dataframe**
```{r}


BigThirteenUniversity <- rbind(BigTwelveUniversity, University_48["Kansas",])
tail(BigThirteenUniversity)

```
**Finding mean of BigThirteenUniverisity**


```{r}

BigThirteen_mean <- colMeans(BigThirteenUniversity)
BigThirteen_mean

```



```{r}
BigThirteen_mean <- append(BigThirteen_mean, "BigThirteen", 0)
BigThirteen_mean
```

**Removing Kansas from University_48**

```{r}
University_46 <- University_47 %>% filter(School != "Kansas")
head(University_46)
```




**Adding BigTwelveMean to University_47**

```{r}
University_47 <- rbind(University_46, BigThirteen_mean)
rownames(University_47) <- NULL
University_47 <- column_to_rownames(University_47, "School")
tail(University_47)

```
 **Typecasting to *double**

```{r}
University_47[ColumnNames] <- sapply(University_47[ColumnNames], as.numeric)
head(University_47)
```

```{r}

maximum_dist <- dist(University_47, method = "maximum")
#maximum_dist
```

School closer to the BigTen

```{r}

maximum_dist <- as.matrix(maximum_dist)

#colnames(maximum_dist)

which.min(maximum_dist["BigThirteen", -47])
min(maximum_dist["BigThirteen",-47])

```


Adding Kentucky to BigThirteenUniveristy dataframe
```{r}


Big14University <- rbind(BigThirteenUniversity, University_47["Kentucky",])
tail(Big14University)

```
**We got our 4th school, kentucky**


\newpage

5. How does the hypothetical 14-team cluster created in (4) compare to the 
actual 14-team Big Ten Conference? For both the hypothetical 14-team Big Ten
Conference and the actual 14-team Big Ten Conference, compute the cluster
centroid, the distance from each cluster member to the cluster centroid,
and average distance between the observations in the cluster.
What do you observe when comparing these two clusters? 
Which cluster has the smaller average distance between observations? Is this
surprising? Explain.



**First take Penn.State,Nebraska,Maryland, and Rutgers to BigTenUniversity and**
**make a actual BigTenUniveristy.**
```{r}


Actual_Big14Univ <- rbind(BigTenUniversity, 
                     University_47[c("Penn.State","Nebraska",
                                     "Maryland","Rutgers"),])
Actual_Big14Univ

```
**Get the mean of hypothetical and actual Big 14 Universities.**

```{r}
ActualBig14_mean <- colMeans(Actual_Big14Univ)
ActualBig14_mean
class(ActualBig14_mean)

```



```{r}
HypotheticalBig14_mean <- colMeans(Big14University)
HypotheticalBig14_mean
class(HypotheticalBig14_mean)
```

**Append ActualBig14_mean and hypotheticalBig14_mean to respective data frames.**


```{r}
Actual_Big14Univ <- rbind(Actual_Big14Univ, ActualBig14_mean)

```

```{r}
rownames(Actual_Big14Univ)[15]<-"Actual_centroid"
Actual_Big14Univ
```


```{r}
Hypothetical_Big14Univ <- rbind(Big14University, HypotheticalBig14_mean)
rownames(Hypothetical_Big14Univ)[15]<-"Hypothetical_centroid"
Hypothetical_Big14Univ
```
**Computing the distance from each cluster member to the centroid of actual**
**big 14 universities.**

```{r}

maximum_dist <- dist(Actual_Big14Univ, method = "euclidian")
#maximum_dist
```


```{r}

maximum_dist <- as.matrix(maximum_dist)
maximum_dist["Actual_centroid",-15]

AverageDistance_Actual <- mean(maximum_dist["Actual_centroid",-15])
AverageDistance_Actual

```


**Computing the distance from each cluster member to the centroid of**
**hypothetical big 14 universities.**

```{r}

maximum_dist <- dist(Hypothetical_Big14Univ, method = "euclidian")
# maximum_dist
```


```{r}

maximum_dist <- as.matrix(maximum_dist)
maximum_dist["Hypothetical_centroid",-15]

AverageDistance <-mean(maximum_dist["Hypothetical_centroid",-15])
AverageDistance
```




## Observations about Actual 14-team cluster and Hypothetical 14-team cluster

### Hypothetical Big Ten

  * Hypothetical Big ten is formed by finding the nearest schools in vector 
    space to the original Big Ten. Missouri, Iowa.State, kansas, and Kentucky 
    are the schools added to the original Big Ten Schools.   
    
  * The centroid of the 14- team cluster is 0.74241125, 0.64506026, -0.63065174
    0.58001582, 0.984887328	with respect to "StadiumCapacity","Latitude",
    "Longitude", "Endowment ($000)","Enrollment".   
    
  * The average distance between the observations in the cluster is 1.773114

  
### Actual Big Ten

  * Actual 14- team Big Ten is formed by adding  "Penn.State","Nebraska",
    "Maryland", "Rutgers" to the original Big Ten schools  during the years
    1990-2014.    
    
  * The centroid of the actual 14 team school is 0.89840478, 0.71634130, 
    -0.16645063,	0.60267129, 1.1694306 with respect to "StadiumCapacity",
    "Latitude", "Longitude", "Endowment ($000)","Enrollment".   
    
  * The average distance between the observatiobs in the cluster is 1.934058.
  
  
### Stadium Capacity

  The stadium capacity of hypothetical cluster is approximatly 0.74 standard 
  deviation away from the mean, whereas the actual cluster is approximatly 0.90 
  standard deviation from the mean, which tells us that the stadium capacity for
  actual 14 team cluster is higher.   
  
  
### Endowment 

  The Endowment of the hypothetical cluster is around 0.58 standard deviation 
  away from the mean and the actual cluster is 0.61 standard deviation away from
  the mean. This also conclude that the actual cluster has more funding than the
  hypothetical.
  
  
### Enrollment

  The hypothetical cluster has a lower enrollment(0.98 standard deviation away
  from the mean) compared to the actual 14-team cluster(1.30 standard deviation
  away from the mean). 
  
  
### Conclusion  
  
  The hypothetical cluster has the smaller average distance(1.77) between 
  observations compared to the actual cluster(1.93). It is surprising to see
  the actual cluster more spread than the hypothetical cluster. It indicates
  that the four schools added between 1990 to 2014 were not the most similar
  schools to the original Big Ten as found in the cluster analysis.
  
  
   



